RCloud: Integrating Exploratory Visualization, Analysis and Deployment

### 2014-03-31 introduction.tex
working

In addition, consider recent trends in interactive visualization
technologies in the open source community. Projects led by Jeff Heer
include Prefuse, Flare, Protovis, d3, and Vega, in that order.
All but the first of these libraries take advantage of web technologies,
and all but the second rely on open standards. Hadley Wickham is the
author of ggplot, ggplot2, and ggvis. ggplot and ggplot2 generate
static graphics, while ggvis provides convenient interactive visualizations
in R, using the R package Shiny and the Vega abstract visualization grammar,
opening a separate web server for each call.

@@

Today's state-of-the-art in environments for data science teams are extremely
powerful, but also fragmented. Instead of creating a new solution that risks
even further fragmentation, we wish to identify and remedy the shortcomings
in current technology for the practice of data science. The following
requirements summarize the main pain points:

@@

\item Technology transfer. In our experience, development
of new analyses and visualizations is done by \emph{hackers}
and \emph{scripters} (using Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}) who are, by definition, familiar
to some extent with programming and software development.
On the other hand, \emph{application
users} are the ones extracting value from the data using the tools
developed by hackers and scripters. Traditionally, the communication
between these two world is done by IT staff. In an environment where
business needs change possibly many times a week, manual IT support
does not scale: there needs to exist
infrastructure to connect the two worlds.
%% The two classes of users
%% have quite different perspectives but we
%% infrastructure for continuous deployment of visualizations.

\item Coexistence. In the current data science ecosystem, the
isolation of current exploratory visualization and data analysis
environments is a threat to wider adoption of modern techniques for
both sides. We specifically see the richness of interactive
visualization tools isolated from the richness of statistical
programming environments.
We wish to provide infrastructure such that
developments on one side can more easily be employed on the other. In
addition, this system itself should be easily accessible
to the rest of the organization.

\item Discoverability. From our experience in working with data
analyst teams, we find that repeated work (``how can I connect to
database X?'', ``how do you get clean data from column Y from feed
Z?'') typically arises from lack of communication of solved
subproblems. We wish to provide transparency about the work of
team members in the organization.

@@

Technologies which fail to address these requirements
are holding back the practice and impact of interactive visual
analysis. In this paper, we discuss the design and deployment of
RCloud, a prototype environment that supports collaborative data
analysis, visualization and web deployment satisfying all requirements
above. We will report on the design decisions, tradeoffs and
limitations, and compare the concepts in RCloud to similar proposals.

RCloud uses the R language~\cite{RCoreTeam:2013:R} as the main provider of
statistical analysis tools. In fact, as we describe in
Section~\todo{where?}, RCloud notebooks are sequences of R snippets
interspersed with Markdown for easy generation of reports. In
Section~\todo{where?}, we describe how RCloud enables
interactive visualization widgets in this setting. In
Section~\ref{sec:limitations}, we discuss some consequences of
designing for a single language, and how this was received by
our (small, but active) user community.

@@

The facilities for \emph{sharing, viewing and embedding} RCloud
notebooks address technology transfer, and are described in
Section~\todo{where?}. RCloud notebooks are edited and shared as web
pages. In order to support the tight integration of R and HTML5
technologies, RCloud uses a full, bi-directional communication channel
between the web browser and a running R process, described in
Section~\todo{where?}.

@@

The requirement of continuous technology transfer creates challenges
about the \emph{stability} of the many moving pieces: if notebooks are
always updated and deployed automatically, how can the system handle
changing requirements and public interfaces? We discuss our choice of
transparent, integrated version control for RCloud notebooks in 
Section~\ref{sec:notebooks}.

>>>>

In addition, consider trends in interactive visualization software
in the open source community. Projects led by Jeff Heer
include Prefuse, Flare, Protovis, d3, and Vega, in that order.
All but the first of these libraries take advantage of web technologies,
and all but the second rely on open standards. Hadley Wickham is the
author of ggplot, ggplot2, and ggvis. ggplot and ggplot2 generate
static graphics, while ggvis provides convenient interactive visualizations
in R, using the R package Shiny and the Vega abstract visualization grammar,
opening a separate web server for each call. These projects point to
the trend toward visual analytics frameworks and systems that are
increasingly interactive, composable, and accessible through the web.

@@

Today's state-of-the-art environments for data science teams are unquestionably
powerful, but also fragmented. Instead of creating a new solution that promotes
even further fragmentation, we attempt to identify and remedy the shortcomings
in the current practice of data science. The following requirements summarize
the main pain points:

@@

\item Technology transfer. In our experience, development
of new analyses and visualizations is done by \emph{hackers}
and \emph{scripters} (using Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}) who are, by definition,
at least somewhat familiar with programming and software development.
On the other hand, \emph{application
users} ultimately derive value from the data by means of the tools
developed by hackers and scripters. Traditionally, the connection
between these two worlds is made by IT staff. In an environment where
business needs can change every day, manual IT support does not scale:
we need to create software infrastructure to connect the two worlds.
(The path backward is equally important: when problems discovered in
production require expert help, we would like the experts to be
able to immediately apply their full range of analytic tools without
moving the problem back to the EDA environment).
%% The two classes of users
%% have quite different perspectives but we
%% infrastructure for continuous deployment of visualizations.

\item Coexistence. In the current data science ecosystem, the
isolation of exploratory visualization and data analysis
environments is a hinderance to wider adoption of modern techniques
on both sides. We specifically see the richness of interactive
visualization tools as still being isolated from the richness
of statistical programming environments.
There is an opportunity to provide a framework so that developments
on either side are more easily employed by the other.
In addition, the integrated results should be more easily accessible
to the rest of the organization.

\item Discoverability. From our experience in working with data
analyst teams, we find that repeated work (``How can I connect to
database X?'', ``How do you get clean data from column Y from feed
Z?'') arises from lack of communication about previously solved subproblems.
We wish to provide transparency about the work of team members in the organization.

@@

Technologies which fail to address these requirements
are holding back the practice and impact of interactive visual
analysis. In this paper, we discuss the design and deployment of
RCloud, a prototype environment that supports collaborative data
analysis, visualization and web deployment satisfying all requirements
above. We will report on the design decisions, tradeoffs and
limitations, and compare the concepts in RCloud to related proposals.

RCloud uses the R language~\cite{RCoreTeam:2013:R} as the core set
of statistical analysis tools. In fact, as we describe in
Section~\ref{sec:notebooks}, RCloud notebooks are sequences of R snippets
interspersed with Markdown for easy generation of reports. In
Section~\ref{sec:visualization}, we describe how RCloud enables
interactive visualization widgets in this setting. In
Section~\ref{sec:limitations}, we discuss some consequences of
designing for a single language, and how this was received by
our (small, but active) user community.

@@

The facilities for \emph{sharing, viewing and embedding} RCloud
notebooks address technology transfer, and are described in
Section~\todo{where?}. RCloud notebooks are edited and shared as web
pages. To support tight integration of R and HTML5, RCloud uses a
full bi-directional communication channel between the web browser
and a running R process, described in Section~\todo{where?}.

@@

The requirement of continuous technology transfer creates challenges
about the \emph{stability} of the many moving pieces: if notebooks are
updated and deployed automatically, how can the system handle
changing requirements and public interfaces? We discuss our choice of
transparent, integrated version control for RCloud notebooks in 
Section~\ref{sec:notebooks}.

### 2014-03-31 system.tex
working

Using GitHub was mainly a tactical decision to reduce our
time-to-prototype. At the same time, it had the beneficial effect of
allowing our backend to communicate with separate processes over HTTP
and JSON. This had the serendipitous advantage of quickly enabling
full search over RCloud notebooks, using Apache SOLR.
%
This perspective has implications for
systems-based research and prototyping in visual analytics:
For small teams to efficiently explore novel ideas toward the
improvement of a large, complex software ecosystem, it pays off
to use existing solutions as far as possible.
%
Even though it involved learning standards that are not directly
related to visual analytics, our early decision to use the GitHub
gist API was succesful. We expect related projects would gain many
of the same benefits by adopting a similar strategy.

>>>>

Using GitHub was mainly a tactical decision to reduce our
time-to-prototype. At the same time, it had the beneficial effect of
allowing our backend to communicate with separate processes over HTTP
and JSON. This had the serendipitous advantage of quickly enabling
full search over RCloud notebooks, using Apache SOLR.
%
This perspective has implications for
systems-based research and prototyping in visual analytics:
For small teams to efficiently explore novel ideas toward the
improvement of a large, complex software ecosystem, it pays off
to use existing solutions as far as possible.
%
Even though it involved learning standards that are not directly
related to visual analytics, our early decision to use the GitHub
gist API was successful. Future projects can gain many
of the same benefits by adopting a similar strategy.
